---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# armacmp

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/dirkschumacher/armacmp.svg?branch=master)](https://travis-ci.org/dirkschumacher/armacmp)
[![Codecov test coverage](https://codecov.io/gh/dirkschumacher/armacmp/branch/master/graph/badge.svg)](https://codecov.io/gh/dirkschumacher/armacmp?branch=master)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/dirkschumacher/armacmp?branch=master&svg=true)](https://ci.appveyor.com/project/dirkschumacher/armacmp)
<!-- badges: end -->

The goal of `armacmp` is to create a DSL to formulate linear algebra code in R that is compiled to C++ using the Armadillo Template Library. It also offers an mathematical optimization that uses `RcppEnsmallen` to optimize functions in C++.

The scope of the package is linear algebra and Armadillo. It is not meant to evolve into a general purpose R to C++ transpiler.

This is currently an *experimental prototype* with most certainly bugs or unexpected behaviour. However I would be happy for any type of feedback, alpha testers, feature requests and potential use cases.

Potential use cases:

* Speed up your code :)
* Quickly estimate `Rcpp` speedup gain for linear algebra code
* Learn how R linear algebra code can be expressed in C++ using `armacmp_compile` and use the code as a starting point for further development.
* Mathematical optimization with `arma_optim`
* ...


## Installation

``` r
remotes::install_github("dirkschumacher/armacmp")
```

## Caveats and limitations

* *speed*: R is already really fast when it comes to linear algebra operations. So simply compiling your code to C++ might not give you a *significant and relevant* speed boost. The best way to check is to measure it yourself and see for your specific use-case, if compiling your code to C++ justifies the additional complexity.
* *NAs*: there is currently no NA handling. In fact everything is assumed to be double (if you use matrices/vectors).
* *numerical stability*: Note that your C++ code might produce different results in certain situations. Always validate before you use it for important applications.

## Example

You can compile R like code to C++. Not all R functions are supported.

```{r}
library(armacmp)
```

Takes a matrix and returns its transpose.

```{r}
trans <- armacmp(function(X) {
  return(t(X))
})
trans(matrix(1:10))
```

Or a slightly larger example using QR decomposition

```{r, echo=TRUE, eval=TRUE}
# from Arnold, T., Kane, M., & Lewis, B. W. (2019). A Computational Approach to Statistical Learning. CRC Press.
lm_cpp <- armacmp(function(X, y = type_colvec()) {
  qr_res <- qr(X)
  qty <- t(qr.Q(qr_res)) %*% y
  beta_hat <- backsolve(qr.R(qr_res), qty)
  return(beta_hat, type = type_colvec())
})

# example from the R docs of lm.fit
n <- 70000 ; p <- 20
X <- matrix(rnorm(n * p), n, p) 
y <- rnorm(n)
all.equal(
  as.numeric(coef(lm.fit(X, y))),
  as.numeric(lm_cpp(X, y))
)
```

## API

`armacmp` always compiles functions. Every function needs to have a `return` statement with an optional type argument.

```{r, eval=FALSE}
my_fun <- armacmp(function(X, y = type_colvec())) {
  return(X %*% y, type = type_colvec())
}
```

A lot of linear algebra functions/operators are defined as well some control flow (for loops and if/else).
Please take a look at the [function reference article](https://dirkschumacher.github.io/armacmp/articles/function-reference.html) for more details what can be expressed.

### Optimization of arbitrary and differentiable functions using `ensmallen`

The package now also supports optimization of functions using `RcppEnsmallen`. Find out more at [ensmallen.org](https://ensmallen.org/).

All code is compiled to C++. During the optimization there is no context switch back to R.

#### Arbitrary function

Here we minimize `2 * norm(x)^2` using simulated annealing.

```{r}
# taken from the docs of ensmallen.org
optimize <- arma_optim(
  data = list(),
  evaluate = function(x) {
    return(2 * norm(x)^2)
  },
  optimizer = optimizer_SA()
)

# should be roughly 0
optimize(matrix(c(1, -1, 1), ncol = 1))
```

Optimizers:

* Simulated Annealing through `optimizer_SA`
* Conventional Neural Evolution `optimizer_CNE`
* ...

#### Differentiable functions

Here solve a linear regression problem using L-BFGS.

```{r}
optimize_lbfgs <- arma_optim(
  data = list(design_matrix = type_matrix(), response = type_colvec()),
  evaluate = function(beta) {
    return(norm(response - design_matrix %*% beta)^2)
  },
  gradient = function(beta) {
    return(-2 %*% t(design_matrix) %*% (response - design_matrix %*% beta))
  },
  optimizer = optimizer_L_BFGS()
)

# this example is taken from the RcppEnsmallen package
# https://github.com/coatless/rcppensmallen/blob/master/src/example-linear-regression-lbfgs.cpp
n <- 1e6
beta <- c(-2, 1.5, 3, 8.2, 6.6)
p <- length(beta)
X <- cbind(1, matrix(rnorm(n), ncol = p - 1))
y <- X %*% beta + rnorm(n / (p - 1))

# Run optimization with lbfgs fullly in C++
optimize_lbfgs(
  design_matrix = X,
  response = y,
  beta = matrix(runif(p), ncol = 1)
)
```

Optimizers:

* L-BFGS through `optimizer_L_BFGS`
* Gradient Descent through `optimizer_GradientDescent`
* ...

### When does `armacmp` improve performance?

It really depends on the use-case and your code. In general Armadillo can combine linear algebra operations. For example the addition of 4 matrices `A + B + C + D` can be done in a single for loop. Armadillo can detect that and generates efficient code. 

So whenever you combine many different operations, `armacmp` _might_ be helpful in speeding things up.

We gather some examples on the wiki to further explore if compiling linear algebra code to C++ actually makes sense for pure speed reasons.

### Related projects

* [nCompiler](https://github.com/nimble-dev/nCompiler) - Code-generate C++ from R. Inspired the approach to compile R functions directly instead of just a code block as in the initial version.

### Contribute

`armacmp` is experimental and has a volatile codebase. The best way to contribute is to write issues/report bugs/propose features and test the package with your specific use-case.

### Code of conduct

Please note that the 'armacmp' project is released with a
[Contributor Code of Conduct](CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.
