---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# armacmp

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/dirkschumacher/armacmp.svg?branch=master)](https://travis-ci.org/dirkschumacher/armacmp)
[![Codecov test coverage](https://codecov.io/gh/dirkschumacher/armacmp/branch/master/graph/badge.svg)](https://codecov.io/gh/dirkschumacher/armacmp?branch=master)
<!-- badges: end -->

The goal of `armacmp` is to create an experimental DSL to formulate linear algebra code in R that is compiled to C++ using the Armadillo Template Library.

Currently just a quick idea and prototype. If this sounds useful, let me know.

## Installation

``` r
remotes::install_github("dirkschumacher/armacmp")
```

## Example

You can compile R like code to C++. Not all R functions are supported.

```{r}
library(armacmp)
```


Takes a matrix and returns its transpose.

```{r}
trans <- armacmp(function(X) {
  return(t(X))
})
trans(matrix(1:10))
```

Equivalent to R's crossprod (`t(x) %*% y`)

```{r example}
# by default arguments are assumed to be matrices
# but you can also make it explicit
crossprod2 <- armacmp(function(X = type_matrix(), Y = type_matrix()) {
  return(t(X) %*% Y)
})
x <- matrix(1:100000)
all.equal(crossprod2(x, x), crossprod(x, x))
microbenchmark::microbenchmark(
  crossprod2(x, x),
  crossprod(x, x),
  t(x) %*% x
)
```

Compute the coefficient of a linear regression problem:

```{r}
# by default X is assumed a matrix
lm_fit <- armacmp(function(X, y = type_colvec()) {
  return(solve(X, y))
})
X <- model.matrix(mpg ~ hp + cyl, data = mtcars)
y <- matrix(mtcars$mpg)
all.equal(as.numeric(qr.solve(X, y)), as.numeric(lm_fit(X, y)))
```

Or a C++ version of plogis:

```{r}
plogis2 <- armacmp(function(x = type_colvec()) {
  return(1 / (1 + exp(-x)))
})
all.equal(as.numeric(plogis2(1:10)), stats::plogis(1:10))
```

Or make predictions in a logistic regression model given the coefficients:

```{r}
log_predict <- armacmp(function(coef = type_colvec(), new_X) {
  res <- new_X %*% coef
  score <- 1 / (1 + exp(-res))
  return(score, type = type_colvec())
})

formula <- I(mpg < 20) ~ -1 + hp + cyl
X <- model.matrix(formula, data = mtcars)
glm_fit <- glm(formula, data = mtcars, family = binomial())

all.equal(
  as.numeric(log_predict(coef(glm_fit), X)),
  as.numeric(predict(glm_fit, newdata = mtcars, type = "response"))
)
```


Forward and backward solve are implemented

```{r}
backsolve2 <- armacmp(function(x, y = type_colvec()) {
  return(backsolve(x, y))
})
forwardsolve2 <- armacmp(function(x, y = type_colvec()) {
  return(forwardsolve(x, y))
})

# example from the docs of backsolve
r <- rbind(
  c(1, 2, 3),
  c(0, 1, 1),
  c(0, 0, 2)
)
x <- matrix(c(8, 4, 2))
all.equal(backsolve(r, x), backsolve2(r, x))
all.equal(forwardsolve(r, x), forwardsolve2(r, x))
```

### For loops

```{r}
for_loop <- armacmp(function(X, offset = type_scalar_numeric()) {
  X_new <- X
  # only seq_len is currently supported
  for (i in seq_len(10 + 10)) {
    # use replace to update an existing variable
    replace(X_new, log(t(X_new) %*% X_new + i + offset))
  }
  return(X_new)
})

for_loop_r <- function(X, offset) {
  X_new <- X
  for (i in seq_len(10 + 10)) {
    X_new <- log(t(X_new) %*% X_new + i + offset)
  }
  return(X_new)
}

all.equal(
  for_loop_r(matrix(as.numeric(1:1000), ncol = 10), offset = 10),
  for_loop(matrix(as.numeric(1:1000), ncol = 10), offset = 10)
)

microbenchmark::microbenchmark(
  for_loop_r(matrix(1:1000, ncol = 10), offset = 10),
  for_loop(matrix(1:1000, ncol = 10), offset = 10)
)
```

### A faster `cumprod`

```{r}
cumprod2 <- armacmp(function(x = type_colvec()) {
  return(cumprod(x))
})

x <- as.numeric(1:1e6)
bench::mark(
  cumprod(x),
  as.numeric(cumprod2(x))
)
```


### Return type

```{r, echo=TRUE, eval=TRUE}
return_type <- armacmp(function(X) {
  return(sum(log(X)), type = type_scalar_numeric())
})
X <- matrix(1:1000, ncol = 10)
all.equal(
  return_type(X),
  sum(log(X))
)
return_type(X)
```

### `colSums` et al.

```{r}
colSums2 <- armacmp(function(X) return(colSums(X)))
rowSums2 <- armacmp(function(X) return(rowSums(X)))
colMeans2 <- armacmp(function(X) return(colMeans(X)))
rowMeans2 <- armacmp(function(X) return(rowMeans(X)))

X <- matrix(1:100, ncol = 10)
all.equal(colSums(X), as.numeric(colSums2(X)))
all.equal(rowSums(X), as.numeric(rowSums2(X)))
all.equal(colMeans(X), as.numeric(colMeans2(X)))
all.equal(rowMeans(X), as.numeric(rowMeans2(X)))
```

### Just compile and look at the C++ code

You can also take a look at the code itself, in case you would like to generate a C++ snippet that you can further extend.

```{r}
armacmp_compile(function(new_X, coef = type_colvec()) {
  res <- new_X %*% coef
  score <- 1 / (1 + exp(-res))
  return(score, type = type_colvec())
}, "log_predict")
```


## API

### Inputs

You can define your inputs using the standard function syntax. By default paramters are of type `type_matrix`. But they can have other types such as:

* `type_colvec` - a column vector
* `type_rowvec` - a row vector
* `type_scalar_integer` - a single int value
* `type_scalar_numeric` - a single double value

### Body

* `<-` you can use assignments that cause a C++ copy. As most operations return armadillo expressions, this is often not a problem. All assignments create new matrix variables. Other types are not possible at the moment :(.
* `...` many more functions :)

### Return

All functions need to return a value using the `return` function.

## TODO

The idea is to implement a number of more complex constructs and also infer some data types to generate better code.

```{r, echo=TRUE, eval=FALSE}
qr_lm_coef <- armacmp(function(X, y) {
  # TBD
  qr_res <- qr(X)
  qty <- t(qr.Q(qr_res)) %*% y
  beta_hat <- backsolve(qr.R(qr_res), qty)
  return(beta_hat)
})
```

```{r, echo=TRUE, eval=FALSE}
if_clause <- armacmp(function(X, y) {
  test <- sum(exp(X)) < 10 # infers that test needs to be bool
  if (test) {
    return(X %*% y + 10)
  } else {
    return(X %*% y)
  }
})
```

### Related projects

* [nCompiler](https://github.com/nimble-dev/nCompiler) - Code-generate C++ from R. Inspired the approach to compile R functions directly.
