---
title: "Function reference"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Function reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(armacmp)
```

This documents describes all language elements that can be compiled to Armadillo flavored C++ code.

## Function signature

`armacmp` always compiles functions. Every function needs to have a `return` statement with an optional type argument.

```{r, eval=FALSE}
my_fun <- compile(function(X, y = type_colvec())) {
  return(X %*% y, type = type_colvec())
}
```

You can define your inputs using the standard function syntax. By default paramters are of type `type_matrix`. But they can have other types such as:

* `type_colvec` - a column vector
* `type_rowvec` - a row vector
* `type_scalar_integer` - a single int value
* `type_scalar_numeric` - a single double value
* `type_scalar_bool` - a single boolean value

All functions need to return a value using the `return` function. The  `return` function can have an optional return type argument with `return(value, type = type_rowvec())`

## Type system

The package supports only a very limited number of types. For scalar values these are `bool`, `int` and `double`. All matrices and vectors are of type `arma::mat/colvec/rowvec` for double matrices, `arma::imat/icolvec/irowvec` for integer matrices and `arma::umat/ucolvec/urowvec` for unsigned integer matrices (used for indicies).

All operations are performed either on Armadillo types or on scalar types and not on R's data structures.

The package tries to figure out the correct types of variables and functions, but you can give type hints. If the package cannot figure out the type, the C++ type `auto` is used which can have [suboptimal performance](http://arma.sourceforge.net/faq.html) when used together with Armadillo operations.

## Operators

### Assignments

Example:

```{r, eval = FALSE}
compile(function(X) {
  X2 <- X + 1
  X2 <- X2 + 1
  return(X2)
})
```

### Multiplication

* `%*%` - *matrix multiplication* translates to the `*` operator on both `arma::mat` types and regular variables/expressions.
* `*` - *elementwise multiplication* gets translated to `%` for `arma::mat` types and to `*` otherwise.

### The rest

The following operators are translated to their C++ counter-part.

* `/`
* `+`
* `-`
* `!`
* `<`
* `>`
* `>=`
* `<=`

## Functions

### General math functions

* `x ^ y`
* `exp`
* `abs`
* `log`
* `sum`
* `sqrt`
* `cos`
* `acos`
* `cosh`
* `acosh`
* `sin`
* `asin`
* `sinh`
* `asinh`
* `tan`
* `atan`
* `tanh`
* `atanh`

### Linear algebra

* `t`
* `chol`
* `diag`
* `QR`
* `solve`
* `ncol`
* `nrow`
* `crossprod`
* `tcrossprod`
* `det`
* `[i, j]`
* `[i]`
* `length`

### Rest

* `cumsum`
* `sort`
* `unique`
* `cumsum`
* `cumprod`
* `pnorm`
* `seq_len` returns a `arma::colvec` (note that is corresponds to a sequence of numerics)
* `a:b` returns an `arma::colvec` from `a` to `b` (note that is corresponds to a sequence of numerics)
* `rep.int` returns a `arma::colvec` (note that is corresponds to a sequence of numerics)

## Control Flow

Currently if/else, for and while loops are supported.

### if/else

```{r, eval=FALSE}
if_clause <- compile(function(X) {
  # infers that test needs to be bool (auto)
  test <- sum(log(X)) < 10
  X_new <- X
  if (test) {
    X_new <- (t(X) %*% X) + 10
  } else if (sum(X) < 10) {
    X_new <- t(X) %*% X
  } else {
    X_new <- (t(X) %*% X) + 10
  }
  return(X_new)
})
```

Just make sure all return statements return the same type.

Also `if/else` cannot be used as an expression:

```{r, eval=FALSE}
# this is not possible
x <- if (y) 1 else 2
```


### For loops

For loops are supported:

```{r, eval=FALSE}
# currently only seq_len is available to create sequences of ints
x <- seq_len(n)
for (i in x) {
  ...
}
```

Example:

```{r, eval=FALSE}
for_loop <- compile(function(X, offset = type_scalar_numeric()) {
  X_new <- X
  # only seq_len is currently supported
  for (i in seq_len(20)) {
    X_new <- log(t(X_new) %*% X_new + i + offset)
  }
  return(X_new)
})
```

### While loops

```{r, eval=FALSE}
x <- 0
while (x < 10) {
  x <- x + 1
}
```

### Functions

You can also define functions within your functions.

```{r}
fun <- compile(function(X) {
  square <- function(y) {
    return(y^2)
  }
  Y <- square(X)
  return(Y)
})
fun(matrix(1:10))
```
