---
title: "Function reference"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Function reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(armacmp)
```

# Function reference

This documents describes all language elements that can be compiled to Armadillo flavored C++ code.

## Function signature

`armacmp` always compiles functions. Every function needs to have a `return` statement with an optional type argument.

```{r, eval=FALSE}
my_fun <- armacmp(function(X, y = type_colvec())) {
  return(X %*% y, type = type_colvec())
}
```

You can define your inputs using the standard function syntax. By default paramters are of type `type_matrix`. But they can have other types such as:

* `type_colvec` - a column vector
* `type_rowvec` - a row vector
* `type_scalar_integer` - a single int value
* `type_scalar_numeric` - a single double value

All functions need to return a value using the `return` function. The  `return` function can have an optional return type argument with `return(value, type = type_rowvec())`

## Operators

### Assignments

Example:

```{r, eval = FALSE}
armacmp(function(X) {
  X2 <- X + 1
  X2 <- X2 + 1
  return(X2)
})
```

### Multiplication

* `%*%` - *matrix multiplication* translates to the `*` operator on both `arma::mat` types and regular variables/expressions.
* `*` - *elementwise multiplication* gets translated to `%` for `arma::mat` types and to `*` otherwise.

### The rest

The following operators are translated to their C++ counter-part.

* `/`
* `+`
* `-`
* `!`
* `<`
* `>`
* `>=`
* `<=`

## Functions

### General math functions

* `x ^ y`
* `exp`
* `abs`
* `log`
* `sum`
* `sqrt`
* `cos`
* `acos`
* `cosh`
* `acosh`
* `sin`
* `asin`
* `sinh`
* `asinh`
* `tan`
* `atan`
* `tanh`
* `atanh`

### Linear algebra

* `t`
* `chol`
* `diag`
* `QR`
* `solve`
* `ncol`
* `nrow`
* `crossprod`
* `tcrossprod`
* `det`
* `[i, j]`
* `[i]`

### Rest

* `cumsum`
* `sort`
* `unique`
* `cumsum`
* `cumprod`
* `pnorm`
* `seq_len` returns a `Rcpp::NumericVector`
* `rep.int` returns a `arma::colvec`

## Control Flow

Currently if/else and for loops are supported.

### if/else

```{r, eval=FALSE}
if_clause <- armacmp(function(X) {
  # infers that test needs to be bool (auto)
  test <- sum(log(X)) < 10
  if (test) {
    return((t(X) %*% X) + 10)
  } else if (sum(X) < 10) {
    return(t(X) %*% X)
  } else {
    return((t(X) %*% X) + 10)
  }
})
```

Just make sure all return statements return the same type.

Also `if/else` cannot be used as an expression:

```{r, eval=FALSE}
# this is not possible
x <- if (y) 1 else 2
```


### For loops

For loops are supported:

```{r, eval=FALSE}
# currently only seq_len is available to create sequences of ints
x <- seq_len(n)
for (i in x) {
  ...
}
```

Example:

```{r, eval=FALSE}
for_loop <- armacmp(function(X, offset = type_scalar_numeric()) {
  X_new <- X
  # only seq_len is currently supported
  for (i in seq_len(20)) {
    X_new <- log(t(X_new) %*% X_new + i + offset)
  }
  return(X_new)
})
```

### While loops

```{r, eval=FALSE}
x <- 0
while (x < 10) {
  x <- x + 1
}
```


### Functions

You can also define functions within your functions.

```{r}
fun <- armacmp(function(X) {
  square <- function(y) {
    return(y^2)
  }
  Y <- square(X)
  return(Y)
})
fun(matrix(1:10))
```
